---
title: "Project 1: Loan Data"
author: "Anthony Morast and Daniel Nix"
date: "February 3, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = FALSE)

  packages <- c("dplyr", "lattice", "plotrix")
  for(package in packages) {
    require(package)
  }
  source("lc.r")
  lcData <- getGoodData("LoanStats3a.csv")
  lcData <- intRateToNumeric(lcData)
  
  par <- getGoodStanding(lcData)
  subpar <- getDefault(lcData)
  
  par$annual_inc <- remove_outliers(par$annual_inc)
  subpar$annual_inc <- remove_outliers(subpar$annual_inc)
  
  lcData <- rbind(par, subpar)
  lcData$loan_status = factor(lcData$loan_status, levels=c("Fully Paid", "Charged Off"))
  lcData$annual_inc <- remove_outliers(lcData$annual_inc)
```

## Overview
The goal of this project is to summarize loan data and search for trends among borrowers. 
A data set publicly available, provided by the Lending Club.
It provides data about loans provided by the peer-to-peer lending company from 2007 to 2011.
It includes information about the the size of loans, lender ratings, home ownership, and many others. 
Data studied in detail is described in the Data Description section.

The dataset is observational data collected for each loan fulfilled using Lending Club. 
The sample population is limited to people using Lending Club and will likely not be representative of the entire population.
This means that conclusions drawn from this dataset should only be trusted when making decisions within Lending Club itself.
Randomization is not applied because the data is a census of the Lending Club loan population. The data we've used, and data 
for other time periods, [can be found on Lending Club](https://www.lendingclub.com/info/download-data.action).



## Data Description
The following data is analyzed in detail.

+ Loan Amount (Quantitative): The amount of a loan is simply the amount in $USD. The loan amount may have some bearing on how likely a person is to pay back. On one hand, a large loan amount may indicate a long time to pay and more time to default. However, this may not be the case.  

+ Grade (Qualitative): Loan grade is a measure by Lending Club used to assign interest rates to lendees. Grades range from A1 with 5.32% interest to G5 with 30.99% interest and depend on the loan application and credit report.

+ Interest Rate (Quantitative): The APR interest rate for each loan. This may correlate with the loan grade.

+ Loan Status (Qualitative): Status is a measure of how loan payment stands. These range from Fully Paid, meaning the entire balance has been paid back, to Charged Off for which "there is no longer a reasonable expectation of further payments". 

+ Employment Length (Quantitative): The length of time each lendee has held their most recent job. Reliability to work may be an indicator of reliability to pay back a loan.

+ Home Ownership (Qualitative): Describes the current status of home ownership for each lendee. The category includes Own, Mortgage, and Rent. 

+ Annual Income (Quantitative): The amount in $USD each lendee makes as an individual. This category does not inculde any income made by a spouse. Joint income is included as a category but was not analyzed in this report.

+ Delinquent 2 Years (Qualitative): This is a measure of how long a loan has been left delinquent. Varying degrees of delinquincy are included as binary categories, but two years delinquent is used in this report to be equivalent to default.

+ Debt to income ratio (Quantitative): Debt to Income ratio (DTI) is a measure of how much a person owes versus annual income. For example, if a person owes $2,000 and has an annual income of $20,000 has a DTI of 10%.

## Summary
```{r}
  # int_rate, loan_amnt, grade, emp_length, home_ownership, annual_inc, delinq_2yrs, dti
  # Note: only 40 rows have a loan status other than "Fully Paid" or "Charged Off" in this dataset

  # par to subpar interest rate summaries
  summary(par$int_rate) 
  summary(subpar$int_rate)
```
  
```{r}
  # par to subpar annual income
  summary(par$annual_inc)
  summary(subpar$annual_inc)
```
  
```{r}
  # par to subpar dti
  summary(as.numeric(as.character(par$dti)))
  summary(as.numeric(as.character(subpar$dti)))
```

Do a data summary looking at mean, sd, etc

## Data Visualization

### Data Overview
The following pairs plots gives a good overview of each feature of the dataset. 
```{r}
  library(lattice)
  # pairs(lcData)
```
From this scatterplot matrix we see much of the data needs to be massaged to derive valuable information. For example, the multiple black boxes for `emp_length` indicates this data needs to be changed prior to any analysis concerning the feature. 

### Brief Analysis of Lending Club Grades
Based on [historic Lending Club data](https://www.lendingclub.com/info/statistics.action), the `grades` of the loans, which are determined by Lending Club, should be good indicators of their quality. The plot below determines this is mostly true. The mosaic plot shows, in general, an 'A' grade loan is much less likely to default whereas loans given the grade 'G' appear to default 3-4 times as often. 
```{r}
  mosaicplot(lcData$loan_status ~ lcData$grade, xlab="Loan Status", ylab="Loan Grade", main="Loan Status vs Grade")
```  

### Interest Rates Compared to Loan Grades
Information pertaining to the likelihood of default for low grade loans begs the question "Who would purchase these subprime investments?" Supposedly, the most attractive feature of low grade loans is their higher interest rate. To verify this we plotted the interest rates of the loans against their grades.    
```{r}
  #grade to interest rate 
  bwplot(lcData$int_rate ~ lcData$grade, ylab="Interest Rate", xlab="Loan Grade", main="Interest Rate vs Grade" )
```

We can see, for the most part, the interest rates increase significantly for lower grade loans. As shown below, the interest rate is largely dependent on the assigned grade of each loan.   
```{r}
  paste("Interest Rate to Grade Correlation: ", cor(lcData$int_rate, as.numeric(lcData$grade)), sep="")
```

### Key Quantitative Variables
Some features that stood out in the data set were `Annual Income` and `Interest Rate`. Our conjecture was lower annual incomes and/or higher interest rates would lead to more defaults. 

Plotting `Annual Income` to `Loan Status` indicates a persons annual income might play a small role in whether or not a borrower would repay a loan through Lending Club. The data shows the mean and median are relatively close to one another, thus there is no drastic difference in the income levels of borrowers that default and those that pay their loans in full. 

```{r}  
  # comparing quantitative variables to loan_status
  # annual income
  bwplot(lcData$loan_status ~ lcData$annual_inc, main="Annual Income to Loan Status", xlab="Annual Income", ylab="Loan Status") 
```

Comparing `Interest Rate` and `Loan Status` produces similar results. Even though our conjecture is somewhat correct, that is "Charged Off" loans typically have higher interst rates than "Fully Paid" loans, the interest rates are not significantly higher, as shown in the chart below. Looking at the summary data for interest rates between the two types of loans, prime and subprime, shows about a 15.5% difference in the average interest rate. 

```{r}
  # int_rate
  bwplot(lcData$loan_status ~ lcData$int_rate, xlab="Interest Rate (%)", ylab="Loan Status", main="Interest Rate vs Loan Status")
```


## Conclusion
Loans are great! Get one today! It's like FREEEEE MONEYYYYY!!!!
